Parallelization of the Probabilistic Roadmap Method with GPU Acceleration

### 2016-07-12 paper/paper.tex
paper anpassungen

\usepackage[ngerman,english]{babel}
 \usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage[applemac]{inputenc}

@@

\usepackage{hyperref}

\usepackage{graphicx}

\usepackage{wrapfig}

@@

\usepackage{listings}
\lstset{breaklines=true}

@@

%\pagestyle{fancy}
%\fancyhead[L]{Seminar \glqq Optimale Steuerung\grqq\\ Prof. Brokate, Prof. Kuttler} %Kopfzeile links
%\fancyhead[C]{} %zentrierte Kopfzeile
%\fancyhead[R]{10. Dezember 2013\\ Manuel Demmeler} %Kopfzeile rechts
\title{ \textmd{\normalsize{High Performance Computing Project}}\\ Parallelization of the Probabilistic Roadmap Method with GPU Acceleration}
\date{}
\author{Manuel Demmeler}

@@

\newcommand{\3}{ ^{3\times3} }
\newcommand{\R}{\mathbb{R}}
\newcommand{\N}{\mathbb{N}}
\newcommand{\M}{\mathbb{M}}
\newcommand{\Sym}{\mathbb{S}^3}
\newcommand{\Mpos}{\M_+^3}
\newcommand{\Orth}{\mathbb{O}^3}
\newcommand{\norm}[1]{\left\lVert#1\right\rVert}
\newcommand{\abs}[1]{\left |#1\right |}
\newcommand{\ub}{\bar{u}}
\newcommand{\yb}{\bar{y}}
\newcommand{\scprod}[2]{\langle#1 , #2 \rangle}

@@

In this project, a parallelized version of the PRM method has been developed in general.
The robotics application then was implemented as an special case of this with GPU support.

@@

The abstract problem treated here is to find a path in a d-dimensional space, which avoids intersecting obstacles. The area occupied by obstacles is given by an indicator function
\begin{equation*}
	I: \Omega \rightarrow \{0, 1\}
\end{equation*}
on a region \( \Omega = [ q_{min,1}, q_{max,1} ] \times ... \times  [ q_{min,d}, q_{max,d} ] \subset \R^d  \). Hence, the goal is to find a continuous curve
\begin{equation*}
	\Gamma: [0,1] \rightarrow \Omega
\end{equation*}
between a given start and endpoint, \(\Gamma(0)=q_b,\ \Gamma(1)=q_e \), 
such that
\begin{equation*}
	I(\Gamma(s))=0 \text{\ f.a.\ } s \in [0,1].
\end{equation*}
The PRM algorithm now grows a graph \(G=(V,E)\) form points \(q_s\) and \(q_e\) by randomly sampling nodes \(v\in \Omega\) and inserting them into $V$.  Two nodes \(v,w \in V\) are connected by an edge, if they are near enough each other and the line between them is free from obstacles. The algorithm terminates, when there exists a path from \(q_s\) to \(q_e\) on G.
\begin{definition}
For two points \(q_1\) and \(q_2\) we define the connection with stepsize \(h>0\) as
	\begin{equation*}
		[q_1,q_2]_h \defgl \{ q = \lambda q_1+(1-\lambda) q_2\ |\ \lambda \in [0,1], \norm{q-q_1} \in \N_0 h \}.
	\end{equation*}
	We say, \(q_1\) and \(q_2\) are connected with stepsize \(h>0\), if
	\(I(q)=0\) for all \(q \in [q_1, q_2]_h\).
\end{definition}

\begin{algorithm} 
	\caption{Probabilistic Roadmap Algorithm}
	\KwData{ \(q_s, q_e\in \Omega, h>0, d_0>0\)}
	\KwResult{ \(q_s=q_1, ... ,q_n=q_e\in \Omega\), 
		such that all \(q_i\) and \(q_{i+1}\) are connected with stepsize \(h\)}
	Initialize graph \(G=(V,E)\) with \(V=\{q_s, q_e\}\) \\
	If \(q_s\) and \(q_e\) are connected, \(E \leftarrow\{ q_s, q_e\}\) \\
	\While{ there exists no path in G between \(q_s\) and \(q_e\)}{
		Sample \(q\in \Omega\) from a probability density  \(p(\ \cdot\ , G)\), resample until \(I(q)=0\) \\
		\For{ all nodes \(\tilde{q} \in V\), with \( \norm{q- \tilde q} \le d_0 \) }{
			\If { $q$ and $\tilde q$ connected with stepsize h} {
				\(E \leftarrow \{q, \tilde{q}\}\)
			}
		}
		\(V \leftarrow{q}\), if at least one edge was inserted \\
	}
	Determine the shortest path \(q_1, ... , q_n\in V\) on \(G\) from \(q_s\) to \(q_e\).
\end{algorithm}

The probability density \(p(\ \cdot\ ,G)\) (depending on the actual graph) can be for example first selecting randomly a node \(q_0 \in V\) and then sampling a new point \(q\) in the neighbourhood of \(q_0\). Theory about success probabilities and convergence results can be found in \cite{prmlec}, \cite{prm1} or \cite{prm2}.

@@

\newpage
\subsection{Project Overview}

Like in the stated robotics example, in many cases the most costly operation of the PRM algorithm are the many evaluations of the indicator function \(I(q)\), 
which however can be done independently for every new \(q\).
Therefore it is implemented in a vectorized way as a function \((q_1, ... , q_P) \mapsto (I(q_1),...,I(q_P))\) by a CUDA kernel, where every GPU thread computes one evaluation.
In the robotics case, this means, the direct kinematic and collision calculation are made on the GPU parallelly for different joint angle configurations.
Although this will the only major application in the project, the PRM solver and the configuration space are implemented separated from each other and connected by an interface to be free to simply add other applications. As an example and for testing, additionally a simple 2D configuration space is provided, 
whose indicator function is read from a black-and-white image.

\begin{figure*}
\begin{center} 
\begin{tikzpicture}[auto]%node distance=0.5cm and 1.8cm]
	\node[block] (prm) {
		\begin{tabular}{c}
		\textbf{PRMSolver}\\
		generate nodes\\
		store graph
		\end{tabular}
	};
	\node [block, right=of prm] (config) {
		\begin{tabular}{c}
		\textbf{Configpace}\\
		indicator fct. \\
		interface
		\end{tabular}
	};
	\node [block, right=of config] (robo) {
		\begin{tabular}{c}
		\textbf{RobotConfigspace}\\
		robot kinematics\\
		collision detection
		\end{tabular}
	};
	\node [block, right=of robo] (gpu) {\textbf{GPU}};

	\path[-stealth,thick, shift left=1.0ex]	(prm) edge (config) ;
	\path[-stealth,thick, shift left=1.0ex]	(config) edge  (prm) ;
	\path[-stealth,thick, shift left=1.0ex]	(config) edge (robo);
	\path[-stealth,thick, shift left=1.0ex]	(robo) edge  (config);
	\path[-,thick]					(robo) edge (gpu);
\end{tikzpicture}
\end{center}
\end{figure*}

@@

@@

The probabilistic roadmap algorithm is applied  to create collision-free trajectories of a robot arm.
For a robot with given joint angles \(q_i\) one can calculate straightforwardly all positions and orientations  of the single parts of the robot.
For this, coordinate frames according to the Denavit-Hartenberg convention are used.
The coordinate frames of the single parts are described by transformation matrices 
$T_i \in R^{4x4}$, which transform points from the body frame into the world frame. They depend on the state $q=(q_1,...,q_d)$ of the robot and on the geometry of the joint angles given by the Denavit-Hartenberg parameters of the robot.

@@

Now the movement of the robot can be described by a curve $q(t) \in \R^d$ in the configuration space. It is collision-free, if at no time, any two polytopes of the robot intersect with each other or with the environment after being transformed into the world frame.
To apply the probabilistic roadmap algorithm, we express this by the indicator function

\begin{equation*}
	I(q)=
	\begin{cases}
		0,	& \text{ for no collision for all given pairs of polytopes } \\
		1,	& \text{ if at least one collision occurs }
	\end{cases}
\end{equation*}

The indicator function is implemented in a vectorized way as an CUDA kernel, which means that, given set of states $q^1,..,q^M$, the indicator function is evaluated for each state by one CUDA thread. Therefore, the kinematics and collision algorithm are implemented as device code for the use in the kernel.

It has to be noted that the trajectories \(q(t)\), the PRM algorithm will generate, are piecewise linear and therefore not differentiable. Hence, they have to be smoothed before the use on a robot. This is not part of the project. The task is only to generate feasible trajectories to be passed to an optimizer.

@@

The Denavit-Hartenberg (DH) convention is a method to describe the geometry of a robots joints, 
that is, how the transformations between the robots coordinate frames look like in dependency of its joint values $q_1,...,q_d$.
A robot consists of $d+1$ bodies each of which has a body fixed coordinate frame $B_i$, which is described by a transformation matrix $T_i$ into the world frame.
It is assumed that each joint is either rotational, that means an rotation around a fixed axis, 
or prismatic, a translation along an axis.
These joints may now be defined through DH parameters $a_{i-1}, \alpha_{i-1}, d_i, \theta_i \in \R$, $i=1,...,d$, 
by setting the relative transformations $T_{i-1,i}=T_{i-1}^{-1} T_{i}$ between two bodies to
\begin{equation}
	T_{i-1,i} = R_z(\theta_i) T_z(d_i) R_x(\alpha_{i-1}) R_x(a_{i-1})
\end{equation}
where $R_x, R_z$ are rotations and $T_x, T_z$ translations around the $x$ and $z$ axes.
With this and $T_0=I$, all transformations $T_i$ are determined.

@@

For the implementation of the kinematics and later the collision algorithm, a geometry library was written, based on the CUDA datatypes float2 for 2D vectors and float4 for 3D vectors.
To guarantee aligned access on the GPU, float4 is used instead of float3. Furthermore, as the geometric calculations made here are numerically not very critical, the usage of double precision has turned out to be not necessary.
The library contains hybrid host and device inline functions for common vector operations such as

\footnotesize
\begin{lstlisting}[language=C++, frame=single, float=*]
__host__ __device__ inline float4& operator += (float4& u, const float4& v);
__host__ __device__ inline float4& operator *=  (float4& u, const float f);
__host__ __device__ inline float4& add(const float4& u, const float4& v, float4& w);
\end{lstlisting}
\normalsize

Operators like $+$ or $*$ are not overloaded to prevent unnecessary temporary variables.

Transformation matrices are implemented as float4 arrays with inline functions for common matrix operations. Again operators with temporaries were avoided, wherever possible.

\footnotesize
\begin{lstlisting}[language=C++, frame=single, float=*]
class trafo4{
  public:
    float4 col[4];
    __host__ __device__ trafo4(){}
    __host__ __device__ trafo4(float a, float alpha, float q, float d);
    __host__ __device__ inline float4& apply(float4& u) const;
    __host__ __device__ inline float4& apply(const float4& u, float4& Tu) const;
    .
    .
    .
}
\end{lstlisting}

The second constructor implements the relative DH transformations.

@@

\footnotesize
\begin{lstlisting}[language=C++, frame=single, float=*]
struct polytope4{
    float4* vertices;	//length n
    int n;
    //edges
    int* dsp;	//length n
    int* cnt;	//length n
    int* dest;	//length m
    int m;
};
\end{lstlisting}

It means, that for $i=0,...,n-1$ the vertices $dest[dsp[i]],...,dest[dsp[i]+cnt[i]-1]$ are the ones connected with vertex $i$ by an edge.
The polytopes used in this project are given by the user as convex hull of a list of vertices. However, to determine the edge graph only from the vertices is a nontrivial task, which is done by Matlab here using provided high level functions.

@@

\footnotesize
\begin{lstlisting}[language=C++, frame=single, float=*]
__host__ __device__ int separating_vector_algorithm(const polytope4& P, const polytope4& Q, const trafo4& tp, const trafo4& tq);
\end{lstlisting}

In our application, the polytopes are mostly parts of the robot, which are given in a body frame and have to be transformed into world frame according to section \ref{dk} first.
However, from the structure of the Chung-Wang algorithm drops out, that mostly only a small part of the polytopes vertices are needed for the computations. Therefore, it would be rather time wasting to transform all vertices before passing them to the algorithm. To avoid this, all polytopes are passed in their body frames together with their transformations and the algorithm transforms a vertex, whenever it occurs in calculations.
This is also the version proposed in \cite{chung}.

@@

The robot configuration space is implemented in the form

\footnotesize
\begin{lstlisting}[language=C++, frame=single, float=*]
template<int ndof>
class RobotConfigspace : public Configspace<ndof>
{
public:
  RobotConfigspace(const Robot<ndof>* robot_,
                   const polytope *polys_,
                   const int* sys_,
                   const int N_,
                   const int *from_, const int *to_,
                   const int M_,
                   const float* mins_, const float* maxs_, const float dq_,
                   const int nbuf_);
  int init(const int ressource_rank=0, const int ressource_size=1);
  int indicator2(const float* qs, const float* qe, int *res, const int N, const int offset);
  .
  .
  .
private:
  const Robot<ndof>* robot; //host object
  Robot<ndof>* robotdev;    //GPU object
  collision4::polytope4data* polydata;    //host object
  collision4::polytope4data* polydatadev; //GPU object
  .
  .
  .
}
\end{lstlisting}

\footnotesize
\begin{lstlisting}[language=C++, frame=single, float=*]
__global__ void kernel_indicator2(const Robot<ndof>* robot,
                                  const collision4::polytope4data* polydata,
                                  const float* qs, int offsets,
                                  const float* qe, int offsete,
                                  int* res,
                                  const int* testpos, const int* testnum,
                                  int N, int numthreads);
\end{lstlisting}

where the constant geometry and robot data is loaded once at the beginning and copied to the GPU by an init function.
The indicator function gets as input a list of border points $q_{start}^j, q_{end}^j, j=0,...,N$ and checks now for each pair, if the indicator function is always 0 on the line in between.

\begin{equation}
	res[j]=\begin{cases}
		0	&\text{if $I(q)=0$ f.a. $q \in \left[qs[j], qe[j]\right]_{\Delta q}$} \\
		1	&else
	\end{cases}
\end{equation}

The border points are stored in structures of arrays.

\begin{equation}
\begin{aligned}
	q_{start}^j &= \left(qs[j],qs[j+\text{offset}],...,qs[j+(d-1)*\text{offset}]\right)^T\\
	q_{end}^j &= \left(qe[j],qe[j+\text{offset}],...,qe[j+(d-1)*\text{offset}]\right)^T
\end{aligned}
\end{equation}

As the PRMSolver only needs to know, which node points can be connected by a line, and not, which intermediate points are free in detail, this version delivers exactly the minimal necessary information. 
Furthermore, it saves CPU and communication time, because the intermediate points on the lines can be directly computed by the GPU threads and do not need to be exchanged.

Algorithms \ref{host} and \ref{kernel} show in pseudo code, how the work is split between host and device.

@@

In order to overlap computation time of GPU and CPU, an asynchronous version of the indicator function was written. Here the function is divided into everything until the kernel launch and a function, which waits, until the kernel has finished and receives the result data.

\begin{lstlisting}
int indicator2_async(const float* qs, const float* qe, int *res, 
					 const int N, const int offset, int &request);
int indicator2_async_wait(int request);
\end{lstlisting}

The implementation also allows overlapping of multiple requests, for example two calls of the launch function and after this the two corresponding waiting calls. 
With this, it is possible to use the full computation time of the GPU.

\newpage
%%%%%%%%%%%%%%%%%%%%%%%%

@@

\begin{lstlisting}
struct block{
	int pos;	 //! position of first vertex
	int num;	 //! current number of vertices stored
	block* next; //! if num==blocksize -> pointer to next block
};
\end{lstlisting}

\begin{lstlisting}
struct graph{
    std::map<int,block*> map; //map for accessing blocks
    std::vector<block> blocks;
    std::vector<float> qstorage; //length ndof*N
    int newblockpos;  //position of next new block
    int blocknum;     //number of used blocks
    .
    .
};
\end{lstlisting}

The key of an block is computed by the mapping
\begin{equation*}
	key(q) = \left\lfloor \frac{q_1}{H} \right\rfloor
\end{equation*}
such that $H>0$ becomes a parameter to adjust the grid density.
This has to be chosen so large that enough nodes are stored in one block in order to have efficient memory access.
On the other side, the smaller it is, the less unnecessary computations have to be done for the neighborhood requests.
For illustration of the structure, the following pseudocode shows the insertion of a node.

@@

For getting the neighbors of a node $q$ it is then enough to look at the keys from $key(q-cD)$ to $key(q+cD)$. For $D=H$, $c=1$ is possible.
The key mapping could be generalized by taking more components of $q$ into account, but that was not done here.

\subsection{Parallelization Models}

@@

	\begin{center}
\begin{tikzpicture}[auto, node distance=0.5cm and 1.8cm]
	\node[block, minimum width=3cm] (prm) {
		\begin{tabular}{c}
		\textbf{PRMSolver 1}
		\end{tabular}
	};

@@

\end{tikzpicture}
\end{center}

@@

\end{tikzpicture}
\end{center}

@@

\newpage

>>>>

\usepackage[english]{babel}
%\usepackage[english,ngerman]{babel}

\usepackage[T1]{fontenc}
%\usepackage{lmodern}
\usepackage[applemac]{inputenc}

\tolerance=250

@@

%\usepackage{hyperref}

%\usepackage{graphicx}
%\usepackage{wrapfig}

@@

\usepackage{listings}
\lstset{breaklines=true}
\renewcommand{\lstlistingname}{Code}

@@

%\pagestyle{fancy}
%\fancyhead[L]{Seminar \glqq Optimale Steuerung\grqq\\ Prof. Brokate, Prof. Kuttler} %Kopfzeile links
%\fancyhead[C]{} %zentrierte Kopfzeile
%\fancyhead[R]{10. Dezember 2013\\ Manuel Demmeler} %Kopfzeile rechts
\title{ %\textmd{\normalsize{High Performance Computing Project}}\\ 
	Parallelization of the Probabilistic Roadmap Method with GPU Acceleration}
\subtitle{High Performance Computing Project}
%\date{}
\author{Manuel Demmeler \and Manfred Liebmann}

\institute{Manuel Demmeler
\at todo ... \\ \email{manuel.demmeler@yahoo.de}
\and
Manfred Liebmann
\at todo ... \\ \email{todo}
}

@@

\newcommand{\R}{\mathbb{R}}
\newcommand{\N}{\mathbb{N}}
\newcommand{\Sym}{\mathbb{S}^3}
\newcommand{\Mpos}{\M_+^3}
\newcommand{\Orth}{\mathbb{O}^3}
\newcommand{\norm}[1]{\left\lVert#1\right\rVert}
\newcommand{\abs}[1]{\left |#1\right |}
\newcommand{\ub}{\bar{u}}
\newcommand{\yb}{\bar{y}}
\newcommand{\scprod}[2]{\langle#1 , #2 \rangle}

@@

In this project, a parallelized version of the PRM method has been developed in general.
A robotics application then was implemented as an special case of this with GPU support.

@@

The abstract problem treated here is to find a path in a d-dimensional space, which avoids intersecting obstacles. The area occupied by obstacles is given by an indicator function
\begin{equation}
	I: \Omega \rightarrow \{0, 1\}
\end{equation}
on a region \( \Omega = [ q_{min,1}, q_{max,1} ] \times ... \times  [ q_{min,d}, q_{max,d} ] \subset \R^d  \). Hence, the goal is to find a continuous curve
\begin{equation}
	\Gamma: [0,1] \rightarrow \Omega
\end{equation}
between a given start and endpoint, \(\Gamma(0)=q_b,\ \Gamma(1)=q_e \), 
such that
\begin{equation}
	I(\Gamma(s))=0 \text{\ f.a.\ } s \in [0,1].
\end{equation}
The PRM algorithm now grows a graph \(G=(V,E)\) form points \(q_s\) and \(q_e\) by randomly sampling nodes \(v\in \Omega\) and inserting them into $V$.  Two nodes \(v,w \in V\) are connected by an edge, if they are near enough each other and the line between them is free from obstacles, based on the following definition \ref{connectiondef}. The algorithm terminates, when there exists a path from \(q_s\) to \(q_e\) on G.

\begin{definition}\label{connectiondef}
For two points \(q_1\) and \(q_2\) we define the connection with stepsize \(h>0\) as
	\begin{equation}
	\begin{aligned}
		[q_1,q_2]_h \defgl \{ &q = \lambda q_1+(1-\lambda) q_2\ | \\ 
				& \lambda \in [0,1], \norm{q-q_1} \in \N_0 h \}.
	\end{aligned}
	\end{equation}
	We say, \(q_1\) and \(q_2\) are connected with stepsize \(h>0\), if
	\(I(q)=0\) for all \(q \in [q_1, q_2]_h\).
\end{definition}

Algorithm \ref{prmalg} shows the sequential version. Hereby, the probability density \(p(\ \cdot\ ,G)\) (depending on the current graph) can be for example first selecting randomly a node \(q_0 \in V\) and then sampling a new point \(q\) in the neighbourhood of \(q_0\). Theory about success probabilities and convergence results can be found in \cite{prmlec}, \cite{prm1} or \cite{prm2}.

\begin{algorithm} \label{prmalg}
	\caption{Probabilistic Roadmap Algorithm}
	\KwData{ \(q_s, q_e\in \Omega, h>0, d_0>0\)}
	\KwResult{ \(q_s=q_1, ... ,q_n=q_e\in \Omega\), 
		such that all \(q_i\) and \(q_{i+1}\) are connected with stepsize \(h\)}
	Initialize graph \(G=(V,E)\) with \(V=\{q_s, q_e\}\) \\
	If \(q_s\) and \(q_e\) are connected, \(E \leftarrow\{ q_s, q_e\}\) \\
	\While{ there exists no path in G between \(q_s\) and \(q_e\)}{
		Sample \(q\in \Omega\) from a probability density  \(p(\ \cdot\ , G)\), resample until \(I(q)=0\) \\
		\For{ all nodes \(\tilde{q} \in V\), with \( \norm{q- \tilde q} \le d_0 \) }{
			\If { $q$ and $\tilde q$ connected with stepsize h} {
				\(E \leftarrow \{q, \tilde{q}\}\)
			}
		}
		\(V \leftarrow{q}\), if at least one edge was inserted \\
	}
	Determine the shortest path \(q_1, ... , q_n\in V\) on \(G\) from \(q_s\) to \(q_e\).
\end{algorithm}

@@

%\newpage
\subsection{Project Overview}

Like in the stated robotics example, in many cases the most costly operation of the PRM algorithm are the many evaluations of the indicator function \(I(q)\), 
which however can be done independently for every new \(q\).
Therefore it is implemented in a vectorized way as a function \((q_1, ... , q_P) \mapsto (I(q_1),...,I(q_P))\) by a CUDA kernel, where every GPU thread computes one evaluation.
In the robotics case, this means, the direct kinematics and collision calculations are made on the GPU parallelly for different joint angle configurations.
Although this will the only major application in the pro\-ject, the PRM solver and the configuration space are implemented separated from each other and connected by an interface to be free to simply add other applications, as shown in figure \ref{interfacefig}. As an example and for testing, additionally a simple 2D configuration space is provided, 
whose indicator function is read from a black-and-white image.

@@

\begin{figure*}
\begin{center} 
\begin{tikzpicture}[auto]%node distance=0.5cm and 1.8cm]
	\node[block] (prm) {
		\begin{tabular}{c}
		\textbf{PRMSolver}\\
		generate nodes\\
		store graph
		\end{tabular}
	};
	\node [block, right=of prm] (config) {
		\begin{tabular}{c}
		\textbf{Configspace}\\
		indicator fct. \\
		interface
		\end{tabular}
	};
	\node [block, right=of config] (robo) {
		\begin{tabular}{c}
		\textbf{RobotConfigspace}\\
		robot kinematics\\
		collision detection
		\end{tabular}
	};
	\node [block, right=of robo] (gpu) {\textbf{GPU}};

	\path[-stealth,thick, shift left=1.0ex]	(prm) edge (config) ;
	\path[-stealth,thick, shift left=1.0ex]	(config) edge  (prm) ;
	\path[-stealth,thick, shift left=1.0ex]	(config) edge (robo);
	\path[-stealth,thick, shift left=1.0ex]	(robo) edge  (config);
	\path[-,thick]					(robo) edge (gpu);
\end{tikzpicture}
\end{center}
\label{interfacefig}
\caption{Project structure. Probabilistic roadmap solver and configuration space implementation are separated.}
\end{figure*}

@@

The probabilistic roadmap algorithm is applied  to create collision-free trajectories of a robot arm.
For a robot with given joint angles \(q_i\) one can calculate straightforwardly all positions and orientations  of the single parts of the robot.
For this, coordinate frames according to the Denavit-Hartenberg convention are used.
The coordinate frames of the single parts are described by transformation matrices 
$T_i \in R^{4 \times 4}$, which transform points from the body frame into the world frame. They depend on the state $q=(q_1,...,q_d)$ of the robot and on the geometry of the joint angles given by the Denavit-Hartenberg parameters of the robot.

@@

Now the movement of the robot can be described by a curve $q(t) \in \R^d$ in the configuration space. It is collision-free, if at no time, any two polytopes of the robot intersect with each other or with the environment after being transformed into the world frame.
To apply the probabilistic roadmap algorithm, we express this by the indicator function
\begin{equation}
	I(q)=
	\begin{cases}
		0,	& \text{ for no collision at any polytope pair} \\
		1,	& \text{ if at least one collision occurs }
	\end{cases}
\end{equation}

The indicator function is implemented in a vectorized way as an CUDA kernel, which means that, given a set of states $q^1,..,q^M$, the indicator function is evaluated for each state by one CUDA thread. Therefore, the kinematics and collision algorithm are implemented as device code for the use in the kernel.

It has to be noted that the trajectories \(q(t)\), the PRM algorithm will generate, are piecewise linear and therefore not differentiable. 
Hence, before the use on a robot they have to be smoothed. 
This is not part of the project. 
The task is only to generate feasible trajectories to be passed to an optimizer.

@@

The Denavit-Hartenberg (DH) convention is a method to describe the geometry of a robots joints, 
that is, how the transformations between the robots coordinate frames look like in dependency of its joint values $q_1,...,q_d$.
A robot consists of $d+1$ bodies each of which has a body fixed coordinate frame $B_i$, which is described by a transformation matrix $T_i$ into the world frame.
It is assumed that each joint is either rotational, that means an rotation around a fixed axis, 
or prismatic, a translation along an axis.
These joints can now be defined through DH parameters $a_{i-1}, \alpha_{i-1}, d_i, \theta_i \in \R$, $i=1,...,d$, 
by setting the relative transformations $T_{i-1,i}=T_{i-1}^{-1} T_{i}$ between two bodies to
\begin{equation}
	T_{i-1,i} = R_z(\theta_i) T_z(d_i) R_x(\alpha_{i-1}) R_x(a_{i-1})
\end{equation}
where $R_x, R_z$ are rotations and $T_x, T_z$ translations aro\-und the $x$ and $z$ axes.
With this and $T_0=I$, all transformations $T_i$ are determined.

@@

For the implementation of the kinematics and later the collision algorithm, a geometry library was written, based on the CUDA datatypes float2 for 2D vectors and float4 for 3D vectors.
To guarantee aligned access on the GPU, float4 is used instead of float3. Furthermore, as the geometric calculations made here are numerically not very critical, the usage of double precision has turned out to be not necessary.
The library contains hybrid host and device inline functions for common vector operations as shown in code \ref{code1}.
Operators like $+$ or $*$ are not overloaded to prevent unnecessary temporary variables.
Transformation matrices are implemented as float4 arrays with inline functions for common matrix operations. Again operators with temporaries were avoided, wherever possible.
The second constructor in code \ref{code2} implements the relative DH transformations.

\footnotesize
\begin{lstlisting}[language=C++, frame=single, float=*, 
	caption=Operator examples of the written float4 library.,
	label=code1
]
__host__ __device__ inline float4& operator += (float4& u, const float4& v);
__host__ __device__ inline float4& operator *=  (float4& u, const float f);
__host__ __device__ inline float4& add(const float4& u, const float4& v, float4& w);
\end{lstlisting}
\normalsize

\footnotesize
\begin{lstlisting}[language=C++, frame=single, float=*,
	caption=Class structure for a geomeric transformation matrix.,
	label=code2
]
class trafo4{
  public:
    float4 col[4];
    __host__ __device__ trafo4(){}
    __host__ __device__ trafo4(float a, float alpha, float q, float d);
    __host__ __device__ inline float4& apply(float4& u) const;
    __host__ __device__ inline float4& apply(const float4& u, float4& Tu) const;
    .
    .
    .
}
\end{lstlisting}
\normalsize

@@

\footnotesize
\begin{lstlisting}[language=C++, frame=single,
	caption=Storing polytopes in CRS format.,
	label=code3
]
struct polytope4{
    //vertices
    float4* vertices;	//length n
    int n; 	
    //edge adjacency matrix
    int* dsp;	//length n
    int* cnt;	//length n
    int* dest;	//length m
    int m;
};
\end{lstlisting}
\normalsize

It means, that for $i=0,...,n-1$ the vertices with indices $dest[dsp[i]],..., dest[dsp[i]+cnt[i]-1]$ are the ones connected with vertex $i$ by an edge.
The polytopes used in this project are given by the user as convex hull of a list of vertices. However, to determine the edge graph only from the vertices is a nontrivial task, which is done by Matlab here using provided high level functions.

@@

In our application, the polytopes are mostly parts of the robot, which are given in a body frame and have to be transformed into world frame according to section \ref{dk} first.
However, from the structure of the Chung-Wang algorithm drops out, that mostly only a small part of the polytopes vertices are needed for the computations. Therefore, it would be rather time wasting to transform all vertices before passing them to the algorithm. To avoid this, all polytopes are passed in their body frames together with their transformations and the algorithm transforms a vertex, whenever it occurs in calculations.
This also is the version proposed in \cite{chung}.

%\footnotesize
%\begin{lstlisting}[language=C++, frame=single, float=*]
%__host__ __device__ int separating_vector_algorithm(const polytope4& P, const polytope4& Q, const trafo4& tp, const trafo4& tq);
%\end{lstlisting}
%\normalsize

@@

\footnotesize
\begin{lstlisting}[language=C++, frame=single, float=*,
	caption=Robot configuration space class structure.,
	label=code4
]
template<int ndof>
class RobotConfigspace : public Configspace<ndof>
{
public:
  RobotConfigspace(const Robot<ndof>* robot_,
                   const polytope *polys_,
                   const int* sys_,
                   const int N_,
                   const int *from_, const int *to_,
                   const int M_,
                   const float* mins_, const float* maxs_, const float dq_,
                   const int nbuf_);
  int init(const int ressource_rank=0, const int ressource_size=1);
  int indicator2(const float* qs, const float* qe, int *res, const int N, const int offset);
  .
  .
  .
private:
  const Robot<ndof>* robot; //host object
  Robot<ndof>* robotdev;    //GPU object
  collision4::polytope4data* polydata;    //host object
  collision4::polytope4data* polydatadev; //GPU object
  .
  .
  .
}
\end{lstlisting}
\normalsize

%\footnotesize
%\begin{lstlisting}[language=C++, frame=single, float=*,
%	caption=GPU collision kernel for testing robot configurations.,
%	label=code5
%]
%__global__ void kernel_indicator2(const Robot<ndof>* robot,
%                                  const collision4::polytope4data* polydata,
%                                  const float* qs, int offsets,
%                                  const float* qe, int offsete,
%                                  int* res,
%                                  const int* testpos, const int* testnum,
%                                  int N, int numthreads);
%\end{lstlisting}
%\normalsize

\footnotesize
\begin{lstlisting}[language=C++, frame=single, float=*,
	caption=GPU collision kernel for testing robot configurations.,
	label=code5
]
__global__ void kernel_indicator2(const Robot<ndof>* __restrict__ robot,
                                  const collision4::polytope4data_restrict polydata,
                                  const float* __restrict__ qs, int offsets,
                                  const float* __restrict__ qe, int offsete,
                                  int* res,
                                  const int* __restrict__ testpos, const int* __restrict__ testnum,
                                  int N, int numthreads)
\end{lstlisting}
\normalsize

%The robot configuration space is implemented in the form
Code \ref{code4} exemplarily shows the implementation of the robot configuration space.
The geometry and robot data is load\-ed once at the beginning and copied to the GPU by an init function.
It is constant during runtime.
The indicator function gets as input a list of border points $q_{start}^j, q_{end}^j, j=0,...,N$ and checks now for each pair, if the indicator function is always 0 on the line in between.
\begin{equation}
	res[j]=\begin{cases}
		0	&\text{if $I(q)=0$ f.a. $q \in \left[qs[j], qe[j]\right]_{\Delta q}$} \\
		%0	&\text{if $I(q)=0$ f.a. $q \in \left[q_{start}^j, q_{end}^j\right]_{\Delta q}$} \\
		1	&else
	\end{cases}
\end{equation}
This computation is done by calling the kernel in Code \ref{code5}.
The border points are stored in structures of arrays.
\begin{equation}
\begin{aligned}
	q_{start}^j = (&qs[j],qs[j+\text{offset}], \\
			&..., qs[j+(d-1)*\text{offset}])^T\\
	q_{end}^j = (&qe[j],qe[j+\text{offset}], \\
			&..., qe[j+(d-1)*\text{offset}])^T
\end{aligned}
\end{equation}
As the PRMSolver only needs to know, which node points can be connected by a line, and not, which intermediate points are free in detail, the kernel delivers exactly this minimal necessary information. 
Furthermore, CPU and communication time is saved,
by computing intermediate points of the lines directly on the GPU and not exchanging them.
Algorithms \ref{host} and \ref{kernel} show in pseudo code, how the work is split between host and device.

@@

Furthermore, in order to overlap computation time of GPU and CPU, an asynchronous version of the indicator function was written. Here the function is divided into everything until the kernel launch and a function, which waits, until the kernel has finished and receives the result data.
This implementation also allows overlapping of multiple requests, for example two calls of the launch function and after this the two corresponding waiting calls. 
With that, it is possible to use the full computation time of the GPU.

%\footnotesize
%\begin{lstlisting}[language=C++, frame=single, float=*,
%	caption=Asynchronous versions of indicator functions.,
%	label=code6
%]
%int indicator2_async(const float* qs, const float* qe, int *res, 
%                     const int N, const int offset, int &request);
%int indicator2_async_wait(int request);
%\end{lstlisting}
%\normalsize

%\newpage
%%%%%%%%%%%%%%%%%%%%%%%%

@@

\footnotesize
\begin{lstlisting}[language=C++, frame=single, float=*,
	caption=Roadmap graph structure.,
	label=code7
]
struct block{
	int pos;	//! position of first vertex
	int num;	//! current number of vertices stored
	block* next; 	//! if num==blocksize -> pointer to next block
};

struct graph{
    std::map<int,block*> map; //map for accessing blocks
    std::vector<block> blocks;
    std::vector<float> qstorage; //length ndof*N
    int newblockpos;  //position of next new block
    int blocknum;     //number of used blocks
    .
    .
};
\end{lstlisting}
\normalsize

The key of an block is computed by the mapping
\begin{equation}
	key(q) = \left\lfloor \frac{q_1}{H} \right\rfloor
\end{equation}
such that $H>0$ becomes a parameter to adjust the grid density.
This has to be chosen so large that enough nodes are stored in one block in order to have efficient memory access.
On the other side, the smaller it is, the less unnecessary computations have to be done for the neighborhood requests.
For illustration of the structure, the following pseudocode shows the insertion of a node.

@@

For getting the neighbors of a node $q$ it is then enough to look at the keys from $key(q_1-cD)$ to $key(q_1+cD)$. For $D=H$, $c=1$ is possible.
The key mapping could be generalized by taking more components of $q$ into account, but that was not done here.

\subsection{Parallelization Approaches}

@@

%\begin{figure}	
\begin{center}
\resizebox{\columnwidth}{!}{
\begin{tikzpicture}[auto, node distance=0.5cm and 1.0cm]
	\node[block, minimum width=3cm] (prm) {
		\begin{tabular}{c}
		\textbf{PRMSolver 1}
		\end{tabular}
	};

@@

\end{tikzpicture}
}
\end{center}
%\caption{First version of parallelization: distributed node generation.}
%\end{figure}

@@

\end{tikzpicture}
}
\end{center}
%\caption{Second version of parallelization: redundant node generation. }
%\end{figure*}

@@

%\newpage
